|                       |
|-----------------------|
| title: "Week 5 Summary Exercises"   |
| output: html_notebook |

5.2.4 Exercises Find all flights that

Had an arrival delay of two or more hours

```{r}
library(tidyverse)
library(nycflights13)
filter(flights, arr_delay >= 120)

```

Flew to Houston (IAH or HOU)

```{r}
filter(flights, dest %in% c("IAH", "HOU"))

```

Were operated by United, American, or Delta

```{r}

filter(flights, carrier %in% c("UA", "AA", "DL"))

```

Departed in summer (July, August, and September)

```{r}
filter(flights, month %in% c(7, 8, 9))

##or

filter(flights, month >= 7, month <= 9)

##or

filter(flights, month %in% 7:9)

```

Arrived more than two hours late, but didn't leave late

```{r}

filter(flights, arr_delay >120, dep_delay <= 0)

```

Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}

filter(flights, dep_delay >=60, dep_delay - arr_delay >30)

```

Departed between midnight and 6am (inclusive)

```{r}

filter(flights, dep_time <= 600| dep_time == 2400 )

```

Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?


```{r}
## a shortcut for x >= left & x <= right,
## yes, one example is
filter(flights, between(month, 7, 9))

##can't do Departed between midnight and 6am (inclusive) as inclusive

```

How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
```{r}
filter(flights, is.na(dep_time))
##cancelled flights

```

Exercise 5.2.4:
Why is NA \^ 0 not missing? Why is NA \| TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA \* 0 is a tricky counterexample!)
```{r}
NA^0

##all values raised to 0 are 1

NA|TRUE

##or TRUE is always TRUE

NA & TRUE

##unknown value

NA * 0

##unknown value

```


5.3.1 Exercises 

How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
arrange(flights, desc(is.na(dep_time)), dep_time)
```

Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
arrange(flights, desc(dep_delay))

arrange(flights, dep_delay)

```

Sort flights to find the fastest (highest speed) flights.

```{r}

mutate(flights, distance/air_time) %>%
  arrange(desc(distance/air_time))

```


Which flights travelled the farthest? Which travelled the shortest?
```{r}

arrange(flights, desc(distance))
arrange(flights, distance)

```


5.7.1 Exercises 
Refer back to the lists of useful mutate and filtering functions. Describe how each operation changes when you combine it with grouping.
```{r}
##group_by does not affect arithmetic functions
##Group_by affects summary functions including: (mean()), offset functions (lead(), lag()), ranking functions (min_rank(), row_number()), operate within each group when used with group_by() in mutate() or filter()
```


Which plane (tailnum) has the worst on-time record?
```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(min_rank(on_time) == 1) %>%
  arrange(desc(n))
```


What time of day should you fly if you want to avoid delays as much as possible?
```{r}

flights %>%
  group_by(hour) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)

```


For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.
```{r}
flights %>%
   filter(arr_delay > 0) %>%
   group_by(dest) %>%
 mutate(
    arr_delay_total = sum(arr_delay),
    arr_delay_prop = arr_delay / arr_delay_total
  ) %>%
  select(dest, month, day, dep_time, carrier, flight,
         arr_delay, arr_delay_prop) %>%
  arrange(dest, desc(arr_delay_prop))
```


Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using lag(), explore how the delay of a flight is related to the delay of the immediately preceding flight.

```{r}
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))

lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 1500, by = 120)) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")

```


Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time of a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?
```{r}
flights %>%
  group_by(dest)%>%
  mutate(speed = distance/air_time) %>%
  select(dest, speed) %>%
  arrange(desc(speed))

```


Find all destinations that are flown by at least two carriers. Use that information to rank the carriers.
```{r}
flights %>%
  group_by(dest) %>%
  mutate(d_carrier = n_distinct(carrier)) %>%
  filter(d_carrier >1) %>%
  group_by(carrier) %>%
   summarize(n_dest = n_distinct(dest)) %>%
   arrange(desc(n_dest))

```


For each plane, count the number of flights before the first delay of greater than 1 hour.
```{r}

flights %>%
  select(tailnum, year, month,day, dep_delay) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(tailnum, year, month, day) %>%
  group_by(tailnum) %>%
  mutate(cumulative_hr_delays = cumsum(dep_delay > 60)) %>%
  summarise(total_flights = sum(cumulative_hr_delays < 1)) %>%
  arrange(desc(total_flights))
  
```


5.6.7 Exercises 

Come up with another approach that will give you the same output as not_cancelled %\>% count(dest) and not_cancelled %\>% count(tailnum, wt = distance) (without using count()).
```{r}
##original version
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled %>% 
  count(dest) 
not_cancelled %>%
  count(tailnum, wt = distance) 

##new version
not_cancelled %>%
 group_by(dest) %>%
  summarise(n = n())

not_cancelled %>%
 group_by(tailnum) %>%
  summarise(n = sum(distance))
```


Our definition of cancelled flights (is.na(dep_delay) \| is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?
```{r}

##most important is dep for departure


```


Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?
```{r}

canceled_per_day <-flights %>%
  mutate(canceled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    canceled_num = sum(canceled),
    flights_num = n(), .groups = "keep"
  )
print (canceled_per_day)
```


Which carrier has the worst delays? 
```{r}

flights %>%
  group_by(carrier) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(arr_delay))

```
```{r}
getwd()

```

